Xolotl: An Intuitive and Approachable Neuron and Network Simulator for Research and Teaching

### 2018-10-25 paper/paper.tex
whitespace

@@

\paragraph{A rich library of network components.} Xolotl comes packaged with hundreds of pre-existing components (compartments, synapses, conductances, and mechanisms) that can be used as building blocks to construct model neurons and networks. ``Compartments'' represent sections of membrane with a single membrane potential, intracellular Calcium concentration, and set of constituent components; and can represent either entire cells or parts of cells. Objects of type ``conductance'' represent populations of ion channels in a compartment that produce transmembrane currents. ``synapse'' objects connect two compartments together by introducing a current in the postsynaptic compartment that depends on the presynaptic compartment's membrane potential. Objects of type ``mechanism'' can represent any intracellular mechanism and can read and modify any other component in the cell, and can run arbitrary dynamical systems within them. Parameters of any of these objects can be easily inspected and modified at any time, either manually or through a programmatic interface.

@@

\paragraph{Limited numerical integration strategies.} Most components in the software are integrated using the exponential Euler method, which has been used in integrating neuronal models (\cite{ohErrorAnalysisSpecialized2006, dayanTheoreticalNeuroscience2001}). However, it may be desirable to use other methods under certain conditions. It is possible to introduce new components that implement other integration schemes, or to modify the integration schemes of existing components, but that requires writing new C++  code. Currently, xolotl can only implement integration schemes with fixed step size.

@@

In this section, we illustrate how xolotl can be used to generate, inspect, and simulate a variety of models. These examples have been chosen to demonstrate various features of xolotl, and are intended to serve as templates upon which researchers and educators can build.

@@

Network models in xolotl consist of compartment objects that can be connected by synapse objects. Two compartments representing different neurons can be connected using synapses using the built-in "connect" method. For example, to connect two single-compartment neurons called "LP" and "PY" using a chemical synapse of type "Cholinergic" with strength 30 nS, we can use

\begin{lstlisting}[style=Matlab-editor]
x.connect('LP','PY','Cholinergic','gbar',30); 
\end{lstlisting}

@@

We have focussed on making our software as easy to use as possible, without sacrificing performance or extensibility. For example, the software and all dependencies can be installed using a single-line installer script from within the MATLAB command line. The installation includes worked example scripts that demonstrate various features of the simulator, that can be run without any configuration.  We decided to built the the front-end interface to xolotl in MATLAB to facilitate interoperability with existing tools for time series analysis, optimization, and parallel computing. This allowed us to build rich tools for visualization and to interact with the simulation.

@@

Xolotl is a purely component-based simulator, and all equations need to be included in a C++ header file that specifies a object. Since our automatic type system binds MATLAB objects to the underlying C++ header files, objects can be inspected and parameters can be modified in the MATLAB workspace. To mitigate some of the drawbacks of the component-oriented paradigm, we have implemented an architecture that allows the user to access the underlying C++ code of any object by simply clicking on the object tree in the MATLAB command line. This feature removes the uncertainty inherent in other component-oriented simulators of the equations underlying each component, and allows the user to modify these equations if needed.

@@

@@

In its current form, xolotl is an efficient and easy to use neuron and network simulator that is actively being used in research. 
Results from an early version of this simulator guided intuition in the modeling of a recently characterized Calcium-dependent Potassium channel found in {\em Drosophila} neuromuscular presynaptic terminals (\cite{bronk2018regulation}). Work on the simulator continues in the open at a publicly accessible repository (https://github.com/sg-s/xolotl/), and the library of conductances, synapses and mechanisms that xolotl ships with grows continuously. While this simulator was intended as a research tool, the many worked examples that are built into it demonstrate how it could also be used as a teaching tool.

@@

\section*{Acknowledgments}
The authors would like to thank Timothy O'Leary and Leandro Alonso for useful discussions, Janis Li for help transcribing many conductance models, and members of the Marder Lab, especially Sonal Kedia, Ekaterina Morozova, Jason Pipkin and Philipp Rosenbaum for careful reading of the manuscript.

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_HH}
	\caption{Simulating a single-compartment Hodgkin-Huxley spiking neuron model. (A) Schematic representation of a single-compartment neuron with three populations of ion channels (colored rectangles). (B) In xolotl, the soma is represented using an object of class ``compartment'' and populations of ion channels are represented by ``conductance'' objects contained within the compartment object. (C) The code snippet shown sets up this neuron model, injects current, integrates and plots the voltage, and displays activation functions, all in a few lines of code. (D) Simulated voltage trace of a Hodgkin-Huxley model with three conductances and 0.2 nA of injected current. Colors indicate the dominant current (gold is fast sodium (NaV), blue is delayed rectifier (Kd), red is Leak). (E) firing rate $vs$. current (f-I) curve of this neuron. (F-G) Steady-state gating functions for activation (m) and inactivation (h) gating variables. (H-I) Voltage-dependence of time constants for activation (m) and inactivation (h) gating variables}
	\label{fig:figurehh}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_manipulate}
	\caption{Manipulating neuron parameters in real time. Any set of parameter in the model can be manipulated; here, the maximal conductance of every conductance type in the model from Fig. \ref{fig:figurehh} is being manipulated using the code snippet shown here. The screenshot shows a GUI with sliders for every parameter of interest that is created by the \texttt{manipulate} method. These sliders can be linked to an arbitrary number of visualization functions. In this example, two visualization functions are used: the built in \texttt{plot} method (A) and a custom function that computes the firing-rate-$vs$.-injected current curve for this neuron (B). Both plots refresh themselves with every movement of any slider, allowing the user to build intuition about how every parameter controls the dynamical behavior of the model. A screen recording of this model being manipulated in real time is included in Supplementary Material}
	\label{fig:figuremanipulate}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_benchmark}
	\caption{Comparison of speed and accuracy of xolotl, NEURON and DynaSim. The top row shows simulations of a tonically-firing Hodgkin-Huxley model with three conductances with constant injected current (as in Fig. \ref{fig:figurehh}). The bottom row shows simulations of  a bursting stomatogastric ganglion neuron model with 8 conductances (as in Fig. \ref{fig:figurenetwork}). Ratio of run-time to simulation time (relative speed) as a function of simulation time step (A, B). Simulation error as a function of the step size (C, D).  Relative speed of integration as a function of the simulation length (E, F). Relative speed of integration, normalized by system size, as a function of the number of compartments simulated simultaneously (G, H). All benchmarks were performed on the same computer, and all simulators using fixed time-step integration methods. NEURON was run using the Python wrapper. }
	\label{fig:figurebenchmark}
\end{figure}

>>>>

\usepackage{tcolorbox}
\tcbuselibrary{breakable}
\usepackage[framemethod=tikz]{mdframed}

@@

\paragraph{A rich library of network components.} Xolotl comes packaged with hundreds of pre-existing components (compartments, synapses, conductances, and mechanisms) that can be used as building blocks to construct model neurons and networks. ``Compartments'' represent sections of membrane with a single membrane potential, intracellular Calcium concentration, and set of constituent components; and can represent either entire cells or parts of cells. Objects of type ``conductance'' represent populations of ion channels in a compartment that produce transmembrane currents. ``synapse'' objects connect two compartments together by introducing a current in the postsynaptic compartment that depends on the presynaptic compartment's membrane potential. Objects of type ``mechanism'' can represent any intracellular mechanism and can read and modify any other component in the cell, and can run arbitrary dynamical systems within them. Parameters of any of these objects can be easily inspected and modified at any time, either manually or through a programmatic interface. \hl{Every compartment exists as a C++ file within the code repository, making it easy to browse these components and edit them. For example, mechanisms are stored under ``c++/mechanism''. }

@@

\begin{tcolorbox}[breakable,notitle,boxrule=0pt,colback=yellow,colframe=yellow]

\paragraph{Limited numerical integration strategies.} Most components in the software are integrated using the exponential Euler method, which has been used in integrating neuronal models (\cite{ohErrorAnalysisSpecialized2006, dayanTheoreticalNeuroscience2001}). Single compartment models may also be integrated using the 4th order Runge-Kutta method. Unbranched multi-compartment models are automatically integrated using an implicit Crank-Nicholson method (\cite{dayanTheoreticalNeuroscience2001}). However, it may be desirable to use other methods under certain conditions. It is possible to introduce new components that implement other integration schemes, or to modify the integration schemes of existing components, but that requires writing new C++  code. Currently, xolotl can only implement integration schemes with fixed step size.
\end{tcolorbox}

@@

In this section, we illustrate how xolotl can be used to generate, inspect, and simulate a variety of models. These examples have been chosen to demonstrate various features of xolotl, and are intended to serve as templates upon which researchers and educators can build. \hl{In every example, we assume that a new xolotl object has been initialized using}

\begin{lstlisting}[style=Matlab-editor]
x = xolotl;
\end{lstlisting}

@@

\begin{tcolorbox}[breakable,notitle,boxrule=0pt,colback=yellow,colframe=yellow]

Network models in xolotl consist of compartment objects that can be connected by synapse objects. Two compartments representing different neurons can be connected using synapses using the built-in "connect" method. For example, in the well-studied pyloric circuit in the crustacean stomatograstic ganglion, a cell called LP synapses onto PY cells using several synapses, that together have an effective combined maximal conductance of around 30 nS. This can be modelled in xolotl by using single compartments to represent the LP and PY cells, and connecting the LP cell to the PY cell using a chemical synapse of type "Glutamatergic" with strength 30 nS using

\end{tcolorbox}

\begin{lstlisting}[style=Matlab-editor]
x.connect('LP','PY','Glutamatergic','gbar',30);
\end{lstlisting}

@@

We have focussed on making our software as easy to use as possible, without sacrificing performance or extensibility. For example, the installation includes worked example scripts that demonstrate various features of the simulator, that can be run without any configuration.  We decided to built the the front-end interface to xolotl in MATLAB to facilitate interoperability with existing tools for time series analysis, optimization, and parallel computing. This allowed us to build rich tools for visualization and to interact with the simulation.

\begin{tcolorbox}[breakable,notitle,boxrule=0pt,colback=yellow,colframe=yellow]

\subsection{The cpplab architecture}

A common theme in the architecture of many simulators and high-performance software is the union of user-facing code written in a scientific programming or user-friendly language and performance-critical code written in a fast, compiled language like C++. The reasons for splitting the codebase across two languages are often pragmatic: user interfaces and second-order tools are often easier to implement in richer programming environments, but performance requirements often mean that code needs to be written in a language like C++ or FORTRAN. While newer projects like Julia aim to merge ease of use and performance in a single language, a number of simulators and software projects embrace this division across languages.  For example, DynaSim, a general purpose dynamical systems simulator, allows for models specified in MATLAB to be compiled into C and executed, speeding up execution (\cite{sherfeyDynaSimMATLABToolbox2018}). NEURON similarly presents a more use-friendly Python interface, while integration is carried out in compiled code (\cite{hinesNEURONSimulationEnvironment1997}). MATLAB's built in "codegen" command allows MATLAB functions to be automatically translated into C code and compiled, which can then be run instead of the MATLAB version. The SWIG project (\cite{beazley1996swig}) offers tools to connect programs written in C or C++ to a variety of scripting languages like Python and Ruby.

\end{tcolorbox}

\begin{tcolorbox}[breakable,notitle,boxrule=0pt,colback=yellow,colframe=yellow]

Our approach in developing "cpplab", our bridge between C++ code and MATLAB, was conceptually different from many of these architectures. Unlike MATLAB's "codegen" function, which assumes the primacy of MATLAB code and makes C code that mirrors the MATLAB code, "cpplab" places C++ code first, and only aims to mimic the structure of objects specified in C++ header files, but not their functionality. Thus, cpplab is able to create a bonafide MATLAB object whose properties reflect that of the C++ object, but not its methods. This simpler task can be easily achieved without the construction of an intermediate interface file, as used in more powerful solutions like SWIG. For the purposes of a neuron and network simulator, this architecture is more robust since only simple parsing of the C++ code is needed. This architecture can be easily used in other applications, and it is trivial to remove some restrictions that the cpplab architecture currently imposes, like lack of support for vectors.

\end{tcolorbox}

@@

Xolotl is a purely component-based simulator, and all equations need to be included in a C++ header file that specifies an object. Since our automatic type system binds MATLAB objects to the underlying C++ header files, objects can be inspected and parameters can be modified in the MATLAB workspace. To mitigate some of the drawbacks of the component-oriented paradigm, we have implemented an architecture that allows the user to access the underlying C++ code of any object by simply clicking on the object tree in the MATLAB command line. This feature removes the uncertainty inherent in other component-oriented simulators of the equations underlying each component, and allows the user to modify these equations if needed.

@@

\begin{tcolorbox}[breakable,notitle,boxrule=0pt,colback=yellow,colframe=yellow]
Finally, we walk through a typical workflow to compare xolotl to other neuron simulators. The first step is to initialize a xolotl object, e.g., using "x = xolotl". This is similar to importing the NEURON package from within Python. Compartments can be created and added to the object tree using "x.add(...)", and channels and mechanisms can be added to these compartments using further "add" commands. This workflow is similar to NEURON, where conductances are ``inserted'' into compartments. In both xolotl and NEURON, the tree of objects can be inspected and individual properties like maximal conductances or reversal potentials can be changed from the command line. The setup phase of the workflow is thus similar to that in NEURON, but differs from equation-oriented environments like Brian or DynaSim, where a set of equations has to be explicitly written out. xolotl differs from NEURON at this stage in that it is possible to ``click through'' the object tree all the way to view the underling C++ code for a particular component, and thus to view the ODE being solved, and the numerical method used to solve it. Since NEURON uses several intermediate files, that may have to be compiled, it is not trivial to identify and view the file that matches an object in the workspace. Another unique feature of xolotl is that every xolotl object is automatically and silently hashed in the background, and that this hash can be inspected at any time. This makes it easy to verify that the object constructed so far is what it should be.
\end{tcolorbox}

@@

In its current form, xolotl is an efficient and easy to use neuron and network simulator that is actively being used in research.
Results from an early version of this simulator guided intuition in the modeling of a recently characterized Calcium-dependent Potassium channel found in {\em Drosophila} neuromuscular presynaptic terminals (\cite{bronk2018regulation}). Work on the simulator continues in the open at a publicly accessible repository (https://github.com/sg-s/xolotl/), and the library of conductances, synapses and mechanisms that xolotl ships with grows continuously. \hl{While this simulator was intended as a research tool, the many worked examples that are built into it demonstrate how it could also be used as a teaching tool. We intend to continue development of this software as we add new features and improve performance. We have developed an automatic testing and deployment framework that routinely tests the publicly available repository and packages a MATLAB toolbox for redistribution. As the user base grows, we intend to transition to finding maintainers who will ensure its continued availability and functionality. }

@@

\section*{Acknowledgments}
\hl{The authors would like to thank Timothy O'Leary and Leandro Alonso for useful discussions, Janis Li for help transcribing many conductance models, and Stephen Van Hooser, Sonal Kedia, Ekaterina Morozova, Jason Pipkin and Philipp Rosenbaum for careful reading of the manuscript.}

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_HH}
	\caption{Simulating a single-compartment Hodgkin-Huxley spiking neuron model. (A) Schematic representation of a single-compartment neuron with three populations of ion channels (colored rectangles). (B) In xolotl, the soma is represented using an object of class ``compartment'' and populations of ion channels are represented by ``conductance'' objects contained within the compartment object. (C) The code snippet shown sets up this neuron model, injects current, integrates and plots the voltage, and displays activation functions, all in a few lines of code. (D) Simulated voltage trace of a Hodgkin-Huxley model with three conductances and 0.2 nA of injected current. Colors indicate the dominant current (gold is fast sodium (NaV), blue is delayed rectifier (Kd), red is Leak). (E) firing rate $vs$. current (f-I) curve of this neuron. (F-G) Steady-state gating functions for activation (m) and inactivation (h) gating variables. (H-I) Voltage-dependence of time constants for activation (m) and inactivation (h) gating variables. \hl{Every panel in this figure except E can be reproduced using built-in methods of xolotl}.}
	\label{fig:figurehh}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_manipulate}
	\caption{Manipulating neuron parameters in real time. Any set of parameter in the model can be manipulated; here, the maximal conductance of every conductance type in the model from Fig. \ref{fig:figurehh} is being manipulated using the code snippet shown here. The screenshot shows a GUI with sliders for every parameter of interest that is created by the \texttt{manipulate} method. These sliders can be linked to an arbitrary number of visualization functions. In this example, two visualization functions are used: the built in \texttt{plot} method (A) and a custom function that computes the firing-rate-$vs$.-injected current curve for this neuron (B). Both plots refresh themselves with every movement of any slider, allowing the user to build intuition about how every parameter controls the dynamical behavior of the model. A screen recording of this model being manipulated in real time is included in Supplementary Material.}
	\label{fig:figuremanipulate}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_benchmark}
	\caption{Comparison of speed and accuracy of xolotl, NEURON and DynaSim. The top row shows simulations of a tonically-firing Hodgkin-Huxley model with three conductances with constant injected current (as in Fig. \ref{fig:figurehh}). The bottom row shows simulations of  a bursting stomatogastric ganglion neuron model with 8 conductances (as in Fig. \ref{fig:figurenetwork}). Ratio of run-time to simulation time (relative speed) as a function of simulation time step (A, B). Simulation error as a function of the step size (C, D).  Relative speed of integration as a function of the simulation length (E, F). Relative speed of integration, normalized by system size, as a function of the number of compartments simulated simultaneously (G, H). All benchmarks were performed on the same computer, and all simulators using fixed time-step integration methods. NEURON was run using the Python wrapper. \hl{Figures similar to this can be generated using the built-in benchmark() method}.}
	\label{fig:figurebenchmark}
\end{figure}
